---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Coplan, Mitchell"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks,"" and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded.  Points assigned to each item appear in this template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets.** An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question:  How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

(1)(a) (2 points) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
table(mydata$CLASS,mydata$RINGS)


```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: The columns consist of ratio data in numerical, nominal, and ordinal data types. The volume data seems to have some outliers. There appears to be roughly equal amounts of male, females, and infants.  There may be potential outliers coming from larger abalones(see Volume, Shuck, and Whole).    

***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
a <- table(mydata$SEX, mydata$CLASS)
a <- addmargins(a)
print(a)
barplot(table(mydata$SEX, mydata$CLASS), col = c("red","blue","green"),beside = T)

```

**Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The graph above displays the frequency of abalones seperated by class and sex. The A1 and A2 class are predominatly cosist of the "I" sex.  The "I" sex is represented in a smaller overall percentage in the A3, A4, and A5 class.The M and F class are roughly in a 1:1 ratio for each class, but make up a greater overall percet of the class in the A3, A4, and A5 class.  ***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
a = sample(1036,200)
work = mydata[a,]
plot(work[, 2:6])
```

-----

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$WHOLE,mydata$VOLUME, col=mydata$CLASS)


```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK, col=mydata$CLASS)
abline(0,max(mydata$SHUCK /mydata$WHOLE))

```

**Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: There appears to be less varibilty(distantance between points) in plot b compared to part a. There also seems to be a positive correleation between WHOLE and SHUCK that can be further statified by age class.  Theere is also a postive correlation between WHOLE and VOLUME. However, this does not appear to have further stratification by class (high variability in class).***

-----

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
a = subset(mydata,SEX  == "I")
b = subset(mydata,SEX  == "F")
c = subset(mydata,SEX  == "M")

hist(a$RATIO,main = paste("Infant Ratio"), col="red")
hist(b$RATIO,main = paste("Female Ratio"), col="blue")
hist(c$RATIO,main = paste("Male Ratio"), col="green")

boxplot(a$RATIO,col="red")
boxplot(b$RATIO,col="blue")
boxplot(c$RATIO,col="green")

qqnorm(a$RATIO,cex=.5,col="red",main = "Infant Ratio")
qqline(a$RATIO,col="black")

qqnorm(b$RATIO,cex=.5,col="blue",main = "Female Ratio")
qqline(b$RATIO,col="black")

qqnorm(c$RATIO,cex=.5,col="green",main = "Male Ratio")
qqline(c$RATIO,col="black")



```

**Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer:  The plots mostly seem to fit the normal distribution, but do reveral a slighly right skew.  This might be due to the outliers per sex.  The outliers can be seen in the boxplots and are greater than the mean, with the exception of one data point in the female sex.  ***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
library(dplyr)
a = subset(mydata,SEX  == "I")
b = subset(mydata,SEX  == "F")
c = subset(mydata,SEX  == "M")
aa=boxplot.stats(a$RATIO)
aa = aa$out
bb=boxplot.stats(b$RATIO)
bb = bb$out
cc=boxplot.stats(c$RATIO)
cc = cc$out

d = c(aa,bb,cc)

e=mydata[which(mydata$RATIO %in% d),]

#mid and extreme outliers
kable(e, format = "html", caption = "RATIO outliers by sex")

library(dplyr)
a = subset(mydata,SEX  == "I")
b = subset(mydata,SEX  == "F")
c = subset(mydata,SEX  == "M")
aa=boxplot.stats(a$RATIO, coef = 3)
aa = aa$out
bb=boxplot.stats(b$RATIO, coef = 3)
bb = bb$out
cc=boxplot.stats(c$RATIO, coef = 3)
cc = cc$out

d = c(aa,bb,cc)

e=mydata[which(mydata$RATIO %in% d),]
print(e)#extreme outliers

```

**Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There are mild outliers in all sexes. There are only 2 extreme outliers, 1 infant and 1 female.  All the outliers are above the mean with the exception of one data point in the female sex.  There is roughly an equal number of outliers for males and females. There are slightly more outliers that are infants.  ***

-----

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow = c(2, 2))
boxplot(mydata$VOLUME~ mydata$CLASS)
title("Volume")
boxplot(mydata$WHOLE~ mydata$CLASS)
title("Whole")
plot(mydata$RINGS, mydata$VOLUME, 
   xlab="RINGS ", ylab="VOLUME", pch=19)
plot(mydata$RINGS, mydata$WHOLE, 
   xlab="RINGS ", ylab="WHOLE", pch=19)
```

**Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: It seems that there is a positve correlation between volume and age. There is also a positve correlation between whole weight and age.  Both Volume and whole weight could therefore be used to help predict the age of the abalone.  However, there appears to a large degree of variablilty in both of these varibles.  Further exploration would be needed to determain how well they correlate and their abiltiy to accureatly predict the age of an abalone. In an ideal world, we would want to be able to further statify this data in order to reduce the variblity. Perhaps if we statify by sex, we will find this data to me more useful.     ***

-----

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
a = aggregate(cbind(VOLUME, SHUCK,RATIO) ~ SEX+CLASS, data = mydata, mean)
a = a[order(a$SEX,a$CLASS),]
b = matrix(c(a[1:3,3],a[4:6,3],a[7:9,3],a[10:12,3],a[13:15,3] ),nrow = 3, ncol = 5,byrow = T, dimnames = list(c("F", "I","M"),
                              c("A1", "A2", "A3","A4","A5")))
c = matrix(c(a[1:3,4],a[4:6,4],a[7:9,4],a[10:12,4],a[13:15,4] ),nrow = 3, ncol = 5,byrow = T, dimnames = list(c("F", "I","M"),
                              c("A1", "A2", "A3","A4","A5")))
d = matrix(c(a[1:3,5],a[4:6,5],a[7:9,5],a[10:12,5],a[13:15,5] ),nrow = 3, ncol = 5,byrow = T, dimnames = list(c("F", "I","M"),
                              c("A1", "A2", "A3","A4","A5")))

kable(b, format = "html", caption = "VOLUME")
kable(c, format = "html", caption = "SHUCK")
kable(d, format = "html", caption = "RATIO")



```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
library(dplyr)
a = aggregate(cbind(VOLUME, SHUCK,RATIO) ~ SEX+CLASS, data = mydata, mean)
av= a %>% select(SEX,CLASS,VOLUME)
as= a %>% select(SEX,CLASS,SHUCK)
ar= a %>% select(SEX,CLASS,RATIO)


# plot
library(ggplot2)
ggplot(data = av, aes(x=av$CLASS, y=av$VOLUME, group = av$SEX)) + geom_line(aes(colour=av$SEX))

ggplot(data = as, aes(x=as$CLASS, y=as$SHUCK, group = as$SEX)) + geom_line(aes(colour=as$SEX))

ggplot(data = ar, aes(x=ar$CLASS, y=ar$RATIO, group = ar$SEX)) + geom_line(aes(colour=ar$SEX))

```

**Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer:  These graphs show clear distinctions can be made between the relationships of abalone, sex, age, ratio, volume, and shuck.  If we attempt to use abolone ratio, volume, or shuck as a predictor of the age, we must also inlcude the sex of the abalone.  There appears to be a positive correlation between age and volume/shuck, but this plateaus(and maybe decreases) at a certain age.  Ratio has a negatvie correlation with age.  Overall, there are clear differnces in these relationships between these varibles when statifying by age.  The main question that I have after viewing this data is, did the orginal researches statitfy by sex when there were attempting to predict the  age of the abalones?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2, 2))
mydata_i= subset(mydata, SEX  == "I" &  RINGS < 16)
mydata_a= subset(mydata, (SEX == "M" | SEX == "F") & RINGS < 16)

#infants volume by ratio
boxplot(mydata_i$VOLUME~ mydata_i$RINGS,ylim = c(0, 1100),col="blue")
#adults volume by ratio
boxplot(mydata_a$VOLUME~ mydata_a$RINGS,ylim = c(0, 1100),col="blue")
#infants Whole weight by ratio
boxplot(mydata_i$WHOLE~ mydata_i$RINGS,ylim = c(0, 400),col="red")
#adults Whole weight by ratio
boxplot(mydata_a$WHOLE~ mydata_a$RINGS,ylim = c(0, 400),col="red")
```

**Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer:   It appears that abalone growth is positively correlated with volume and weight until a certain point where the weight and volume plataue.  This plateau time seems to occur around 10 rings for and effects both infants and adults. In gerneral, it appears that the infants have a higher weights and volumnes compared to the adults at the same ring point.    ***

-----

**Conclusions**

Please respond to each of the following questions (10 points total):

**Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: The first plausible reason to explain the failure of the orginal study may be that there is too much variblity in the sample.  In other words, the predicive features had large distributions and could not be used to find a model that "fit" well and resulted in accurate predictions. The outliers found in question 3b contributed to the high degree of variblity that was observed. Physical measurements can be used to predict the age of the abalone, the acurracey of that prediction would require further exploration.  Whole weight, sex, and volume seem to hold some predictive power.  Abalone growth would also be affected by the genetics and environment in which the abalone was raised.  In order to truly gain predicive power, those inputs would need to be measured as well.  Additionally, the orginal study may not have statifyed by sex, we know from question 5b that this correlates with differences in the age. ***

**Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer:  I would want to know much more about the data.  Mainly, I would be interested in learing if there was a way to seperate the data out by cohorts (ie sex, color).  Even though there may be no differnce when a population is looked at as a whole, there may be relationships between varibles when broken down into cohorts(ie sex color).  I would also want to know the variblity of the sample data. If the sample data is contains high variblity, more data may be needed to tease out relationships.***

**Question 3)  (2 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: A common pitfall is to say that observational is equal to causation.  This is not necessarily the case.  Observational studies can only demonstrate differences that were found between groups and shed light on correlations. In observational studies there may be underlying factors that affect the observed relationship. Observational studies can not determine causation. Observational studies can be used to screen for potential causes, but further studies would need to be done in order to determine causation.  These studies would need to include a control group and reject the null hyothesis.  ***
