---
title: "Data Analysis #2 Version 2 (75 points total)"
author: "Coplan, Mitchell"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

### Instructions

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks", and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in the template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets**. An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question: How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
library(rockchalk)
library(tidyverse)
# library(rockchalk) # base R code replaces requirement for this package
library(car)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```


###Section 1 (5 points)
(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}
mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO  <- mydata$SHUCK / mydata$VOLUME
s  <- rockchalk::skewness(mydata$RATIO)
k  <- rockchalk::kurtosis(mydata$RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$RATIO)
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
hist(mydata$RATIO, col = "blue", xlab = "Ratio",
     main = "Histogram of Ratio")
qqnorm(mydata$RATIO, main = "Q-Q Plot",
       col = ifelse(mydata$RATIO %in% c(boxplot.stats(mydata$RATIO)$out), 
                    "red", "blue"))
qqline(mydata$RATIO)
txt <- c("skewness:", round(s, 2), 
         "kurtosis:", round(k, 2), 
         "excess kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}
mydata$L_RATIO <- log10(mydata$RATIO)
s  <- rockchalk::skewness(mydata$L_RATIO)
k  <- rockchalk::kurtosis(mydata$L_RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$L_RATIO)
par(mfrow = c(1, 3), oma = c(0, 0, 2, 0))
hist(mydata$L_RATIO, col = "blue", xlab = "Log Ratio",
     main = "Histogram of Log Ratio")
qqnorm(mydata$L_RATIO, main = "Q-Q Plot",
       col = ifelse(mydata$L_RATIO %in% c(boxplot.stats(mydata$L_RATIO)$out), 
                    "red", "blue"))
qqline(mydata$L_RATIO)
boxplot(mydata$L_RATIO ~ mydata$CLASS,
        main = "Boxplots of Log Ratio by Class",
        ylab = "Frequency",
        xlab = "Log Ratio Distribution by Class", 
        beside = TRUE,
        col = 'blue',
        pars = list(outcol = "red"))
txt <- c("skewness:", round(s, 2), 
         "kurtosis:", round(k, 2), 
         "excess kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
x <- bartlett.test(RATIO ~ CLASS, data = mydata)
y <- bartlett.test(L_RATIO ~ CLASS, data = mydata)
print(list('ratio:', x, 'log ratio:', y))

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: From the graphs and statistics above we know that RATIO is right skewed.  The data for RATIO does not appear to be normally distributed.  Additionally, the Barlett's test for homogenetiy of variance among RATIO classes, tells us that the variance is not homogeneous between the different classes.  The histogram of L_RATIO suggests this data is normally distributed. Additonally, the Barlett's test for homogenetiy of variance amoung L_RATIO CLASS's, tells us that the variance is homogeneous. The L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes. This is a result of the central limit theorem.  ***

###Section 2 (10 points)
(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
summary(aov(L_RATIO ~ CLASS*SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))


```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: For L_RATIO, the interaction term CLASS:SEX is not significant, with a p-value of 0.87. However when CLASS and SEX are looked at independently, there is a significant difference with p-values of 2e-16 and .001, respectively.  The non-significant interaction term suggests that when the L_RATIO is looked at for all distinct groups of class/sex pairs, there is no difference in mean between the groups. ***


(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: All the CLASS pair comparisons seem to be significantly different, besides the A1-A2 class. The diffenence in ratio seems to increase as the age differece increases (ie diff A1-A2 < A1-A5).  We can see this by looking at the diff value and smaller associated p-value. This is corroborated by looking at the boxplot in section 1b where the downward trend as age increases is visible. The non-significant p-value of 0.94 comparing Males to Females and significant results when comparing F-I and M-I does suggest that males and females can be combined into a single category labeled adults.  For L_RATIO there appears to be no differnece between males and females, only between adults and adolescents.***


###Section 3 (10 points)
(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}
par(mfrow = c(1, 2))
hist(mydata$VOLUME[mydata$TYPE == 'I'], col = "blue", xlab = "Volume",
     main = "Infant Volumes", ylim = c(0, 150))
hist(mydata$VOLUME[mydata$TYPE == 'ADULT'], col = "red", xlab = "Volume",
     main = "Adult Volumes", ylim = c(0, 150))
```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The volumes have different distributions.  The Infants have a right skewed volume, there are more infants that have smaller volumes (not surprising). The adults have a roughly normal distribution of volume.  The majority of adults have a volume >300 while the opposite is true of infants.  Although volume alone could not be used to determine if an abalone is an infant or adult, it may provide us some useful information. ***


(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}

mydata <- mydata %>% mutate(L_SHUCK = log10(SHUCK), L_VOLUME = log10(VOLUME))
a=qplot(mydata$VOLUME,mydata$SHUCK, col=mydata$CLASS)
b=qplot(mydata$L_VOLUME ,mydata$L_SHUCK, col=mydata$CLASS)

c=qplot(mydata$VOLUME,mydata$SHUCK, col=mydata$TYPE)
d=qplot(mydata$L_VOLUME ,mydata$L_SHUCK, col=mydata$TYPE)
grid.arrange(a, b,c,d, nrow = 2)
```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: There is a a lot of overlap in the measured volumne and shuck plots which make it difficult to distinguish any grouping by class or type.  The log-transformations appear to decrease the variability present in the plot and allow for visible grouping of class and type. The infants group together in the lower left corner, while the adults group together in the upper right corner. The classes can also be distinguished with lower classes towards the bottom left corner and higher classes towards the upper right corner. Howerver, the classes of A4 and A5 are difficult to differentiate becuase both are in the upper right hand corner of the graph. The implications of log-transformation for linear regression analysis will be a "better fit" (higher R-squared value) becuase it reduces the variability.  ***


###Section 4 (5 points)
(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The levelcoefficient estimates of class get larger(more negative) as the class increases from 1 to 5. In other words, as the class increases from 1 to 5 the L_SHUCK negative correlation strengthens. This is consistent of our ealier finding in part 2b and the boxplot visible in part 1b ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE is less important than most of the classes and much less important than L_VOLUME. The estimated coefficient of -0.02, this is very close to zero and much smaller when the absolute value is comparted to the other coefficients. This suggests that TYPE may not help in predicting L_SHUCK.  ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

### Section 5 (5 points)
(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}
s  <- rockchalk::skewness(model$residuals)
k  <- rockchalk::kurtosis(model$residuals, excess = FALSE)
k2 <- rockchalk::kurtosis(model$residuals)
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
hist(model$residuals, col = "blue", xlab = "Residuals",
     main = "Histogram of Residuals")
qqnorm(model$residuals, main = "Q-Q Plot",
       col = ifelse(model$residuals %in% c(boxplot.stats(model$residuals)$out), 
                    "red", "blue"))
qqline(model$residuals)
txt <- c("skewness:", round(s, 2), 
         "kurtosis:", round(k, 2), 
         "excess kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}
mydata$RESIDUALS <- model$residuals
gridExtra::grid.arrange(
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + theme_bw() + 
    geom_point(aes(fill = CLASS), shape = 23),
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + theme_bw() + 
    geom_point(aes(fill = TYPE), shape = 23),
  ggplot(mydata, aes(CLASS, RESIDUALS)) + theme_bw() +
    geom_boxplot(aes(fill = CLASS)),
  ggplot(mydata, aes(TYPE, RESIDUALS)) + theme_bw() +
    geom_boxplot(aes(fill = TYPE))
)
bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The calcualtions and displays reveal the residuals of the model have a normal distribution overall and are evenly distributed when catagorized by class and type.  Additonally, Bartlett test of homogeneity of variances tells us the variance of residuals by class is homogeneous.  With this information, it seems that the model does 'fit' well.***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

###Section 6 (5 points)
(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"


max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}
ggplot() + theme_bw() +
  geom_line(aes(volume.value, prop.infants), color = "red") +
  geom_line(aes(volume.value, prop.adults), color = "blue") + 
  geom_vline(xintercept = split.infants) +
  annotate("text", label = paste(round(split.infants, 1), "\nInfants"), 
           x = split.infants + 57, y = 0.43, color = "red") +
  geom_vline(xintercept = split.adults) +
  annotate("text", label = paste(round(split.adults, 1), "\nAdults"), 
           x = split.adults + 57, y = 0.43, color = "blue") +
  geom_hline(yintercept = 0.5) +
  labs(x = "VOLUME", y = "Proportion", 
       title = "Proportion of Adults and Infants Protected")

```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The values suggest that an ideal split will be somewhere between these two values. We want to find the volume point that maximizes the percent of adults harvested and minimized the percent of infants harvested.   ***


-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###Section 7 (10 points)
(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
change <- ((1 - prop.adults) - (1 - prop.infants))
ggplot() + theme_bw() +
  geom_line(aes(volume.value, change)) +
  geom_vline(xintercept = split.infants) +
  annotate("text", label = paste('infant volume =', round(split.infants, 1)), 
           x = split.infants + 20, y = 0.2, angle = 90) +
  geom_vline(xintercept = split.adults) +
  annotate("text", label = paste('adult volume =', round(split.adults, 1)), 
           x = split.adults + 20, y = 0.2, angle = 90) +
  labs(x = "VOLUME", y = "Difference in Proportions Harvested", 
       title = "Difference in Harvest Proportions")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}
smoothed_cut <- volume.value[which.max(smooth.difference)]
ggplot() + theme_bw() +
  geom_line(aes(volume.value, change), color = "red") +
  geom_line(aes(volume.value, smooth.difference), color = "blue") +
  geom_vline(xintercept = volume.value[smoothed_cut]) +
  annotate("text", label = paste('volume =', round(smoothed_cut, 3)), 
           x = smoothed_cut + 20, y = 0.35, angle = 90) +
  labs(x = "Volume", y = "Difference in Proportions Harvested", 
       title = "Difference in Harvest Proportions")

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
(1 - prop.adults)[which.max(smooth.difference)]  # true positive rate
(1 - prop.infants)[which.max(smooth.difference)]  # false positive rate



```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

###Section 8 (10 points)
(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
a1_cutoff=volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786
print(a1_cutoff)# cutoff for no a1 infants
(1 - prop.adults)[which(volume.value == a1_cutoff)] #True positive rate
(1 - prop.infants)[which(volume.value == a1_cutoff)]#False positive rate

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
cutoff=volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391
print(cutoff)
(1 - prop.adults)[which(volume.value == cutoff)] #True positive rate
(1 - prop.infants)[which(volume.value == cutoff)]#False positive rate



```


###Section 9 (5 points)
(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}
cuts <- which(volume.value %in% c(smoothed_cut, a1_cutoff, cutoff))
ggplot(mapping = aes((1 - prop.infants), (1 - prop.adults))) + 
  geom_line(color = 'blue', size = 1) + theme_bw() +
  geom_abline(intercept = 0, slope = 1, color = 'blue', linetype = 2) +
  geom_point(mapping = aes((1 - prop.infants[cuts]), (1 - prop.adults[cuts])),
             shape = 21, size = 3) +
  annotate("text", label = paste('equal harvest/conserve \nvol =', round(cutoff, 1)), 
           x = 0.16, y = 0.83, size = 3) +
  annotate("text", label = paste('zero A1 inf. \nvol =', round(a1_cutoff, 1)), 
           x = 0.35, y = 0.8, size = 3) +
  annotate("text", label = paste('max. difference \nvol =', round(smoothed_cut, 1)), 
           x = 0.25, y = 0.68, size = 3) +
  labs(title = "ROC curve of adult and infant harvest proportions",
       x = "Infant harvest proportion", y = "Adult harvest proportion") + 
  theme(axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5))

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
paste("Area under ROC curve:", 
      flux::auc(x = (1 - prop.infants), y = (1 - prop.adults)))

```


###Section 10 (10 points)
(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	

true_pr <- (1 - prop.adults)[cuts]
false_pr <- (1 - prop.infants)[cuts]

yld <- (true_pr * total.adults + false_pr * total.infants) /
  (total.adults + total.infants)
tribble(
  ~strategy, ~volume, ~true_pr, ~false_pr, ~prop_yield,
  "max difference", smoothed_cut, true_pr[3], false_pr[3], yld[3],
  "zero A1 infants", a1_cutoff, true_pr[1], false_pr[1], yld[1],
  "equal error", cutoff, true_pr[2], false_pr[2], yld[2]
)
```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: Each of the three cutoffs offer their own advantages and disadvantages. Depending on what is most important to the harvesters, they could select one of these three cutoffs.  In order to increase the true positve rate, you must "sacrifice" your false positive rate.  The Zero A1 infants cutoff gives the highest true positive rate, the highest false positive rate, and the highest proportion yield. The max differnece cutoff gives the lowest true posive, the lowest false postive and the lowist yield. And, not surprisingly, the equal error cutoff gives the middle true positive, false positive, and proportion yield. ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?

***Answer:I would outline various choices and explain the implications of each choice. For example, I could explain that by picking the lowewest volume choice (206.786), we would be harvesting the highest amount of adult abalones and have the highest yeild. However, I would explain that in doing so, we would also be harvesting more infants which could have affects on next years harvests or other implications. Or they could take a more conservative choice and harvest at a higher cutoff(262.143), and although the harvest would be smaller, it would increase the likelihood of a successful long term harvesting plan.*** 

2. What qualifications or limitations would you present regarding your analysis?

***Answer:There is always a risk of data quality. The model is only as good as the data it was trained on. Additionally, there were a few outliers in this dataset which suggests that there could be mislabeling.***


3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?

***Answer:If I had to make a suggestion with only the current analysis, I would recommend the conservative choice(262) in order to prevent overharvesting.***

4)  What suggestions would you have for planning future abalone studies of this type? 

***Answer: I would be interesting in ways to decrease the number of false positives and increase the true positive rate. I think one opportunity would be to include an additional feautre in the cutoff point. For example, maybe the color of the abalone could be used to predict age. I would research additional features that could be measured that might be predicive of the abalone's age.  The best model will likely include multiple features. ***